---
title: "5 Ways to Calculate Student Persistence in R"
author: "Christian Million"
date: '2019-07-19'
categories:
draft: no
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
lastmod: '2019-07-17T21:16:52-07:00'
projects: []
slug: 5-ways-to-calculate-student-persistence
subtitle: ''
summary: ''
tags:
- ir
- studentdata
- education
authors: []
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      messages = FALSE)
```


# Motivation

Before I spent the time to learn R, I did all of my analyses in Excel. For most tasks, Excel worked well, but I dreaded calculating student persistence. It was tedious, error prone, and time consuming. Worse yet, I was unable to calculate persistence across many terms, so I had to repeat the process over and over. There was probably a better way for me to do it within Excel, but I never took the time to learn it.

Once I became comfortable enough with R's syntax, I was quickly able to calculate persistence on the fly. It was so exciting that I took the time to convert it into my first function.

I soon found out there are many approaches to calculating persistence. In this post, I'll briefly consider the limitations and benefits of each method.

# What is Student Persistence?

The word 'persistence' can mean many different things to those in academia, depending on the context. For my purposes, I am interested in "term-to-term" persistence, which measures whether a student in Term A subsequently enrolled in Term B. Term-to-term persistence can mean:

  - "Fall-to-Spring" (e.g., a student enrolled in the Fall semester of 2018 and then enrolled in the next Spring semester of 2019)
  
  - "Fall-to-Fall" (e.g., a student enrolled in the Fall semester of 2018 and then enrolled in the next Fall semester of 2019)
  
  - any other combination of terms in sequence -- winter quarters, summer sessions, etc. The word 'term' is used as an institutionally agnostic word for "semester", "quarter", "trimester", or what-have-you.

### How is it used?

The persistence rate of a cohort of students is sometimes used as a metric to measure student success, student progression, or as a feature in a model to predict enrollment for next year. The idea is that if a student returns to next semester, they are more likely to be successful in achieving their academic goals.

### Why is it important?

From an institutional perspective, a college is likely to get more funding when the persistence rate is high. Enrollments increase and more students are likely to earn their degree / certificate (and at a faster rate). The numbers as an institution look better with high persistence. It is important to measure persistence rates for subpopulations, so that if the college implements an intervention, it can measure whether or not that intervention effected the likelihood of those students returning.

From a student perspective, in general, persistence is beneficial because they are making progress towards their academic goal. More units and courses completed implies they will recieve their award / certificate sooner or transfer to another institution.

More persistence isn't always better, though. Career Technical Education (CTE) focused students are a common example of this. Students take classes in one term to gain skills that employers are looking for. Once they gain those skills, and maybe get hired, there is no reason to come back the next term to pursue a terminal certificate or degree or transfer. Returning may mean working less, spending more on tuition, accumulating unnecessary units, or anything not conducive to their goals.

# General Approaches to Calculating Persistence

The title of this post is a little misleading. While I *will* show 5 different ways to calculate student persistence, two of them use a similar *approach*, but a different technical implementation. In general though, I've found these approaches useful:

  1. Joins - Joins 2 data.frames. One contains the initial term, the other contains the subsequent term.
  
  2. Lead - Arranges a data.frame with all term data, then implements a `lead` function to see whether a student ended up in the lead term.
  
  3. Lookup - Similar to Excels VLOOKUP (and Joins, really). Lookup student ID's from initial term in target term.
  
All of these approaches have an important common feature: They create a persistence indicator variable first on the original data, and then separately, we calculate the persistence rate from that indicator. This is crucial for my work, because I need to look at persistence among many subpopulations. The indicator allows more dynamic analysis. Additionally, I sometimes need to export the resulting data into another software (Like PowerBI), and indicator is just one feature of many.

This post should probably be titled "3 Approaches to Calculating a Persistence Indicator" instead.

# Read in Data 

First, we'll read in some student data to use in the following examples. 

```{r}
library(readr, warn.conflicts = FALSE)
library(dplyr, warn.conflicts = FALSE)

# Fake Student Data
url_data <- "https://raw.githubusercontent.com/christian-million/student_data/master/student_data.csv"

# Read in student data
student_data <- read_csv(url_data, col_types = "cnccn")
```

## Approach #1: Joins

In general, this approach takes a `data.frame` containing only records for the initial term and joins against it a `data.frame` containing only the target subsequent term records. In the resulting `data.frame`, if there is a record for the initial term *and* the target subsequent term columns, a student has persisted.

I think the benefit to this approach is that it is intuitive. It's beneficial to use with just 1 term-to-term calculation.

Some downsides are that you have to do some processing before if you don't want to have an unmanageable amount of columns. And your output is a `data.frame` that is structured differently from your input. Also, doing more than one term-to-term (i.e, fall-to-spring-to-fall) is harder and the resulting `data.frame`, again,  would have too many columns.

#### Base R

```{r}
# Create Indicator

  # Create 2 dataframes.
  data_fall_2017 <- student_data[student_data$term == 20177,]
  data_fall_2018 <- student_data[student_data$term == 20187,]
  
  data_fall_2018$persist <- 1
  
  # Join
  data_merged <- merge(data_fall_2017, data_fall_2018, by = c("student_id"), all.x = TRUE)
  
# Calculate
  
  sum(data_merged$persist, na.rm = TRUE)/length(data_merged$persist)
```

#### dplyr

```{r}

# Create Indicator

  # Alternate way of creating the data
  data_fall_2017 <- student_data %>%
    filter(term == "20177")
  
  data_fall_2018 <- student_data %>%
    filter(term == "20187") %>%
    select(student_id) %>%
    mutate(persist = 1)
  
  # Join
  data_merged <- data_fall_2017 %>%
    left_join(data_fall_2018, by = 'student_id')
  
# Calculate
  data_merged %>%
    summarise(persist_rate = sum(persist, na.rm = T)/n())
```

## Approach #2: Lead

In general, this approach takes a single dataframe, ranks the terms, sorts the dataframe by id and term, and then implements a 'lead' style function that finds the next value in the vector. If the `current_term` + 1 is equal to the `next_term` for a given student, then persistence!

The benefit to this approach is that all of the data stays in a single data.frame and you can find the persistence for many terms-to-terms.

A downside to this approach is that it is harder to understand how the calculation is being made, in my opinion. It takes a little more code to implement, too. Though that isn't as much of a problem if you wrap it up in a function.

#### Base R

```{r}
  fall_to_fall <- student_data[student_data$term == 20177 | student_data$term == 20187,]
# Create Indicator

  id <- fall_to_fall$student_id
  term <- fall_to_fall$term

  # Get dense rank of each term
  r <- rank(term, na.last = "keep")
  denseRank <- match(r, sort(unique(r)))
  
  # Dataframe with order
  dd <- data.frame(id, term, denseRank, order = 1:length(id), stringsAsFactors = FALSE)
  
  # Order by ID, then ranked term
  dd <- dd[with(dd, order(id, denseRank)), ]
  
  # For each id, what would be the next ranked term?
  dd$nextrank <- with(dd, ave(denseRank, id, FUN = function(x) c(tail(x, -1), rep(NA, 1))))
  
  # Does each id have a record for the next ranked term?
  dd$persist <- with(dd, ifelse(!is.na(nextrank) & nextrank == denseRank + 1, 1, 0))
  
  # Return to original order.
  fall_to_fall$persist <- dd$persist[with(dd, order(order))]

# Calculate
  
  sum(fall_to_fall$persist[fall_to_fall$term == 20177])/length(fall_to_fall$persist[fall_to_fall$term == 20177])
  
```

#### dplyr

```{r}
  fall_to_fall <- student_data[student_data$term == 20177 | student_data$term == 20187,]
# Create indicator 
  fall_to_fall <- fall_to_fall %>%
    mutate(denseRank = dense_rank(term)) %>% 
    group_by(student_id) %>%
    arrange(denseRank) %>%
    mutate(nextrank = lead(denseRank)) %>%
    mutate(persistence_indicator = case_when(nextrank == denseRank + 1 ~ 1, TRUE ~ 0)) %>%
    select(-nextrank, -denseRank)

# Calculate
  fall_to_fall %>%
    group_by(term) %>%
    summarise(persistence = sum(persistence_indicator)/n())
```

## Approach #3: Lookup

This approach is a cleaner, but more limiting version of the join approach. It came from a function I created that served as an Excel VLOOKUP equivalent.

```{r}
# Using dataframes from above (data_fall_2017 and data_fall_2018)

# Create indicator
  out <- data_fall_2018$persist[match(data_fall_2017$student_id, data_fall_2018$student_id)]
  data_fall_2017$persist <- ifelse(is.na(out), 0, out)
  
# Calculate
  sum(data_fall_2017$persist)/length(data_fall_2017$persist)
```

# Conclusions

In general, I prefer the base R version of Approach #2. This is because, once wrapped into a function, I think it produces the most versatile output with the least amount of code. You could wrap the `dplyr` version of Approach #2 into a function (which I have), but I'm still not as comfortable with Non-Standard Evaluation, though it is nice not to have to reference the data source twice.

Let me know if you know other ways of approaching persistence!

# Appendix

Here are some of the Approaches, wrapped in functions.

#### Approach #2 - Base R

```r
calc_persist <- function (id, term) {
    r <- rank(term, na.last = "keep")
    denseRank <- match(r, sort(unique(r)))
    dd <- data.frame(id, term, denseRank, order = 1:length(id), stringsAsFactors = FALSE)
    dd <- dd[with(dd, order(id, denseRank)), ]
    dd$nextrank <- with(dd, ave(denseRank, id, FUN = function(x) c(tail(x, -1), rep(NA, 1))))
    dd$persist <- with(dd, ifelse(!is.na(nextrank) & nextrank == denseRank + 1, 1, 0))
    dd <- dd[with(dd, order(order)), ]
    return(dd$persist)
}
```
#### Approach #2 - dplyr

```r
calc_persist2 <- function (df, id, rank) {

    enq_id <- enquo(id)
    enq_rank <- enquo(rank)

    df %>% ungroup() %>% 
    mutate(denseRank = dense_rank(UQ(enq_rank))) %>% 
    group_by(UQ(enq_id)) %>%
    arrange(denseRank) %>%
    mutate(nextrank = lead(denseRank)) %>%
    mutate(persist = case_when(nextrank == denseRank + 1 ~ 1, TRUE ~ 0)) %>%
    ungroup() %>%
    select(-nextrank, -denseRank)
}
```
#### Approach #3

```r
vlookup <- function (x, data, key, value, na = 0) {
    m <- match(x, data[[key]])
    out <- data[[value]][m]
    ifelse(is.na(out), na, out)
}
```

